Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-09-19 manuscript/usses_water_ms.tex
adds most of JBT edits; working on PBA edits; updates Fig 3 probability sign

\corrauthor[1]{Andrew T. Tredennick}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}
\author[2]{Andrew R. Kleinhesselink}

\author[3]{Bret Taylor}

@@

\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, Utah 84322}
\affil[2]{Department of Ecology and Evolutionary Biology, University of
California, Los Angeles, Los Angeles, California 90095}
\affil[3]{United States Department of Agriculture, Agriculture Research Station,
U.S. Sheep Experiment Station, Dubois, Idaho 83423}

@@

% 
\begin{abstract}
\textbf{Background.} Precipitation is predicted to become more variable
in the western United States, meaning years of above and below average
precipitation will become more common. Periods of extreme precipitation
are major drivers of ecosystem functioning in water limited grasslands,
but how ecosystems respond to precipitation may change as the duration
of above and below average periods increases. Changes in ecosystem
functional response could reflect compensatory changes in species
composition or species reaching physiological thresholds at extreme
precipitation levels.

\textbf{Methods.} We conducted a five year precipitation manipulation
experiment in a sagebrush steppe ecosystem in Idaho, United States. We
used drought and irrigation treatments (approximately 50\%
decrease/increase) to investigate whether ecosystem functional response
remains consistent at extreme precipitation levels. We recorded data on
aboveground net primary productivity (ANPP), species abundance, and soil
moisture. We fit a generalized linear mixed effects model to determine
if the relationship between ANPP and soil moisture differed among
treatments. We used nonmetric multidimensional scaling to quantify
community composition over the five years.

\textbf{Results.} Ecosystem functional response, defined as the
relationship between soil moisture and ANPP was similar among drought
and control treatments, but the irrigation treatment had a lower slope
than the control treatment. However, ANPP response to availabile soil
moisture was weak and uncertain regardless of treatment, with all slopes
overlapping zero. There was also large variation in ANPP within-years.
Plant community composition was remarkably stable over the course of the
experiment and did not differ among treatments.

\textbf{Discussion.} Despite some evidence that ecosystem functional
response was less positive under extreme wet conditions, the response of
ANPP to soil moisture was consistently weak and community composition
was remarkably stable. Ecosystem functional responses across treatments
was not related to compensatory shifts at the plant community level, but
instead reflects the weak and variable responses of individual species.
Such responses may be due to bet hedging strategies where species
respond weakly to even extreme levels of soil moisture to ensure low
variance of long-term success.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

The functional response of aboveground net primary productivity (ANPP)
to water availability (e.g., soil moisture) can be characterized by
fitting a model to historical observations of ANPP and soil moisture at
a given site. However, the fitted functional response may provide an
incomplete picture because future conditions are likely to be outside
the historical range of variability \citep{Smith2011}. For example,
historical trends may underestimate the potential for the soil
moisture-ANPP relationship to saturate or weaken if soil moisture is
pushed far beyond typical levels. Saturating relationships are actually
common \citep{Hsu2012, Gherardi2015a}, perhaps because other resources,
like nitrogen, become more limiting in wet years than dry years. Knowing
the curvature of the soil moisture-ANPP relationship at extreme
precipitation levels is critical for understanding how ecosystems will
respond to chronic alterations in water availability.

Another problem with relying on historical ecosystem functional
responses is that they are not static. Changes in species identities and
abundances can alter an ecosystem's functional response to water
availability because different species have different physiological
thresholds for producing biomass. \citet{Smith2009} introduced the
`Hierarchical Response Framework' for understanding the interplay of
community composition and ecosystem functioning in response to resource
manipulations over time. In the near term, ecosystem functioning such as
ANPP will reflect the physiological responses of individual species to
the manipulated resource level. For example, ANPP may decline under
simulated drought because the initial community consisted of
drought-intolerant species \citep{Hoover2014}. Ecosystem functioning may
recover over longer time spans as new species colonize or initial
species reorder in relative abundance. For example, ANPP may initially
decline, but eventually rise back to pre-treatment levels once
drought-tolerant species become more abundant and compensate for
drought-intolerant species \citep{Hoover2014}. It is also possible that
ecosystem functioning shifts to a new mean state, reflecting the suite
of species in the new community \citep{Knapp2012}.

Manipulating potentially limiting resources, like precipitation, offers
a route to understanding how ecosystems will respond to resource levels
that fall outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. We propose four alternative predictions for the effect of
precipitation manipulation on the ecosystem functional response to soil
moisture based on the Hierarchical Response Framework (Fig. 1). We
define `ecosystem functional response' as the relationship between
available soil moisture and ANPP. The four predictions are based on
possible outcomes at the community (e.g., community composition) and
ecosystem (e.g., soil moisture-ANPP regression) levels.

First, altered precipitation changes neither ecosystem functional
response nor community composition (Fig. 1, top left). In this case,
changes in ANPP simply follow the soil moisture-ANPP relationship under
ambient conditions. This corresponds to the early phases of the
Hierarchical Response Framework, where ecosystem response follows the
physiological responses of individual species. Second, the ecosystem
functional response changes but community composition remains the same
(Fig. 1, top right). A saturating soil moisture-ANPP response fits this
scenario, where individual species hit physiological thresholds or are
limited by some other resource. Third, the ecosystem functional response
is consistent but underlying community composition changes (Fig. 1,
bottom left). In this case, changes in species' identities or abundances
occur in response to altered precipitation levels and species more
suited to the new conditions compensate for reduced function of initial
species. Fourth, and last, both ecosystem functional response and
community composition change (Fig. 1, bottom right). New species, or
newly abundant species, with different physiological responses
completely reshape the ecosystem functional response.

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ
\citep{Smith2009, Wilcox2016, Knapp2017}. Thus, our task is not to test
the validity of the Hierarchical Response Framework, but rather to amass
information on how quickly ecosystem functional responses change in
different ecosystems. We also need to understand whether changes at the
ecosystem level are driven by community level changes or individual
level responses.

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (approximately \(\pm50\%\)) and
measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the
relationship between available soil moisture and ANPP, and if community
dynamics underlie the ecosystem responses. In particular, we are
interested in the consistency of the soil moisture-ANPP relationship
among treatments. Is the relationship steeper under the drought
treatment, at low soil moisture? Does the relationship saturate under
the irrigation treatment, at high soil moisture? To answer these
questions we fit a generalized linear mixed effects model to test
whether the regressions differed among treatments. We also analyzed
community composition over time, allowing us to place our experimental
results within the framework our competing predictions (Fig. 1).

@@

We conducted our precipitation manipulation experiment in a sagebrush
steppe community at the United States Sheep Experimental Station (USSES)
near Dubois, Idaho (44.2\(^{\circ}\) N, 112.1\(^{\circ}\) W), 1500 m
above sea level. The plant community is dominated by the shrub
\emph{Artemesia tripartita} and three perennial bunchgrasses,
\emph{Pseudoroegneria spicata}, \emph{Poa secunda}, and
\emph{Hesperostipa comata}. During the period of our experiment (2011 --
2015), average mean annual precipitation was 265 mm
year\(\phantom{}^{-1}\) and mean monthly temperature ranged from
-5.2\(^{\circ}\)C in January to 21.8\(^{\circ}\)C in July. Between 1926
and 1932, range scientists at the USSES established 26 permanent 1
m\(^2\) quadrats to track vegetation change over time. In 2007, we
(well, one of us {[}P. Adler{]}) relocated 14 of the original quadrats,
six of which were inside a large, permanent livestock exclosure. We use
these six plots as control plots that have received no treatment, just
ambient precipitation, in the experiment described below.

@@

In spring 2011, we (well, two of us {[}A. Kleinhesselink and P.
Adler{]}) established 16 new 1 m\(^2\) plots located in the same
exclosure as the six control plots. We avoided areas on steep hill
slopes, areas with greater than 20\% cover of bare rock, and areas with
greater than 10\% cover of the shrubs \emph{Purshia tridentata} and/or
\emph{Amelanchier utahensis}. We established the new plots in pairs and
randomly assigned each plot in a pair to receive a ``drought'' or
``irrigation'' treatment.

Drought and irrigation treatments were designed to decrease and increase
the amount of ambient precipitation by 50\%. To achieve this, we used a
system of rain-out shelters and automatic irrigation
\citep{Gherardi2013}. The rain-out shelters consisted of transparent
acrylic shingles 1-1.5 m above the ground that covered an area of
\(2.5\times2\) m. The shingles intercepted approximately 50\% of
incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on
to the adjacent irrigation plot via two suspended sprinklers. Pumping
was triggered by float switches once water levels reached about 20
liters. We disconnected the irrigation pumps each fall and reconnected
them, often with difficulty, each spring. The rain-out shelters remained
in place throughout the year.

We monitored soil moisture in four of the drought-irrigation pairs using
Decagon Devices (Pullman, Washington) 5TM and EC-5 soil moisture
sensors. We installed four sensors around the edges of each plot, two at
5 cm soil depth and two at 25 cm soil depth. We also installed four
sensors in areas nearby the four selected plot pairs to measure ambient
soil moisture at the same depths. Soil moisture measurements were
automatically logged every four hours. We coupled this temporally
intensive soil moisture sampling with spatially extensive readings taken
at six points within all 16 plots and associated ambient measurement
areas. These snapshot data were collected on 06-06-2012, 04-29-2015,
05-07-2015, 06-09-2015, and
05-10-2016\footnote{Dates formatted as: mm-dd-yyyy.} using a handheld
EC-5 sensor.

Analyzing the response to experimental treatments was complicated by the
fact that we did not directly monitor soil moisture in each plot on each
day of the experiment. Only a subset of plots were equipped with soil
moisture sensors, and within those plots, one or more of the sensors
frequently failed to collect data. To remedy these problems, and to
produce average daily soil moisture values for the ambient, drought, and
irrigation conditions, we used a statistical model to predict the
average treatment effects on soil moisture during the course of the
experiment.

@@

We then modeled the daily deviation from ambient conditions of the
drought and irrigation treatments using a linear mixed effects model
with independent variables for treatment, season (winter, spring,
summer, fall), rainfall, and all two-way interactions. Rainy days were
defined as any day in which precipitation was recorded and average
temperature was above 3\(^{\circ}\)C. The day immediately following
rainfall was also classified as rainy. We fit the model using the
\texttt{lme4::lmer()} function \citep{Bates2015} in R \citep{R2016},
with random effects for plot group and date. We weighted observations by
the number of unique sensors or spot measurements that were taken in
each plot on that day. We then used the model to predict the average
daily soil moisture in the treated plots based on the average daily
ambient soil moisture. We could only predict soil moisture in the
treated plots on days for which we took at least one ambient soil
moisture measurement.

@@

We estimated aboveground net primary productivity (ANPP) using a
radiometer to relate ground reflectance to plant biomass \citep[see][
for a review]{Byrne2011}. We recorded ground reflectance at four
wavelengths, two associated with red reflectance (626 nm and 652 nm) and
two associated with near-infrared reflectance (875 nm and 859 nm). At
each plot in each year, we took four readings of ground reflectances at
the above wavelengths. We also took readings in 12 (2015), 15 (2012,
2013, 2014), or 16 (2016) calibration plots adjacent to the experimental
site, in which we harvested all aboveground biomass, dried it to a
constant weight at 60\(^{\circ}\)C, and weighed it to estimate ANPP.

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS and AVHRR bands for NDVI. We
regressed the greenness index against the dry biomass weight from the
ten calibration plots to convert the greenness index to ANPP. We fit
regressions to a MODIS-based index and an AVHRR-based index for each
year and retained the regression with the better fit based on \(R^2\)
values. We then predicted ANPP using the best regression equation for
each year (Appendix 1).

Species composition data came from two sources: yearly census maps for
each plot made using a pantograph \citep{Hill1920} and yearly counts of
annual species in each plot. We extracted the mapped information to
obtain data on the density of all annuals and perennials forbs, the
basal cover of perennial grasses, and the canopy cover of shrubs. We
made a large plot-treatment-year by species matrix, where columns were
filled with either basal cover or density, depending on the measurement
made for the particular species. We standardized the values in each
column so we could analyze the different types of data together. This
puts all abundance values on the same scale, meaning that common and
rare species are weighted equally. This approach is anti-conservative
for detecting community change if we assume that rare species will
respond more than common ones, which biases our approach toward
detecting compositional changes.

@@

\noindent{}where \(\mu_{i}\) is the deterministic prediction from the
regression model for observation \emph{i}, which is associated with plot
\emph{j} and treatment year \emph{t}. \(\boldsymbol{\beta}\) is the
vector of coefficients for the fixed effects in the design matrix
\(\textbf{X}\). Each row of the design matrix represents a single
observation (\(\textbf{x}_i\)) and is a vector with the following
elements: 1 for the intercept, a binary 0 or 1 if the treatment is
``drought'', a binary 0 or 1 if the treatment is ``irrigation'', the
scaled value of VWC, binary ``drought'' value times VWC, and binary
``irrigation'' value times VWC. Thus, our model treats ``control''
observations as the main treatment and then estimates intercept and
slope offsets for the ``drought'' and ``irrigation'' treatments. In
reference to our model, the hypotheses we wish to test are:

@@

@@

We used nonmetric multidimensional scaling (NMDS) based on Bray-Curtis
distances to see if community composition differed among treatments
through time. We first calculated Bray-Curtis distances among all plots
for each year of the experiment and then extracted those distances for
use in the NMDS. Some values of standardized species' abundances were
negative, which is not allowed for calculating Bray-Curtis distances. We
simply added `2' to each abundance value to ensure all values were
greater than zero. We plotted the first two axes of NMDS scores to see
if community composition overlapped, or not, among treatments in each
year. We used the \texttt{vegan::metaMDS()} function \citep{Oksanen2016}
to calculate Bray-Curtis distances and then to run the NMDS analysis. We
used the \texttt{vegan::adonis()} function \citep{Oksanen2016} to
perform permutational multivariate analysis of variance to test whether
treatment plots formed distinct groupings. To test whether treatment
plots were equally dispersed, or not, we used the
\texttt{vegan::betadisper()} function \citep{Oksanen2016}.

@@

Three of our five treatment years fell in years of below average
rainfall (Fig. 2A). Thus, those three years represent a lower magnitude
of absolute change in precipitation experienced by the treatments. ANPP
varied from a minimum of 74.5 g m\(^{-2}\) in 2014 to a maximum of 237.1
g m\(^{-2}\) in 2016 when averaged across treatments (Fig. 2C). ANPP was
slightly higher in irrigation plots and slightly lower in drought plots
(Fig. 2C), corresponding to estimated soil volumetric water content
(VWC) differences among treatments (Fig. 2B). Such differences in soil
VWC indicate our treatment infrastructure was successful. ANPP was
highly variable within years (Fig. 2C).

@@

Ecosystem response to precipitation extremes depends on the
physiological responses of constituent species and the rate at which
community composition shifts to favor species better able to take
advantage of, or cope with, new resource levels \citep{Smith2009}.
Previous work has shown that community compositional shifts can be both
rapid, on the order of years \citep{Hoover2014}, and slow, on order of
decades \citep{Knapp2012, Wilcox2016}. Thus, a lingering question is how
the time scales of ecosystem response and community change differ among
ecosystems, which can be answered by manipulating precipitation to reach
extreme levels.

The results of our five year experiment in a sagebrush steppe conform to
two of our four predictions, depending on treatment. Neither ecosystem
functional response nor community composition changed under chronic
drought (Fig. 3A, Fig. 4), consistent with the top left prediction in
Fig. 1. Ecosystem functional response under chronic irrigation was
different from the control treatment, but community composition remained
unchanged (Fig. 3A, Fig. 4), consistent with the top right prediction in
Fig. 1. The altered ecosystem functional response under irrigation
matched our expectations because we found evidence for a saturating
response for the irrigation treatment (Fig. 3A, right panel). However,
the response of ANPP to soil moisture in all treatments was consistently
weak (Table 1, Fig. A2-2) and within-year variation of ANPP was large
(Fig. 3B)

@@

First, it could be that our drought manipulation was not large enough to
induce a response. That is, a 50\% decrease in any given year may not be
abnormal given our site's historical range of variability
\citep{Knapp2017}. We cannot definitively rule out this possibility, but
we have reason to believe our drought treatment \emph{should} have been
large enough. Using the methods described by \citet{Lemoine2016}, we
calculated the percent reduction and increase of mean growing season
precipitation necessary to reach the 1\% and 99\% extremes of the
historical precipitation regime at our site (Fig. A3-3). The 1\%
quantile of precipitation at our site is 110 mm, a 47\% reduction from
the mean, and the 99\% quantile is 414 mm, a 77\% increase from mean
growing season precipitation (Appendix 3). Thus, our drought treatment
represented extreme precipitation amounts, especially in years where
ambient precipitation was below average (Fig. 2A). The irrigation
treatment may have been too small, yet that is the treatment where we
did observe an effect (Fig. 3).

>>>>

\corrauthor[1]{Andrew T. Tredennick}{\href{mailto:atredenn@gmail.com}{\nolinkurl{atredenn@gmail.com}}}
\author[1,2]{Andrew R. Kleinhesselink}

\author[3]{J. Bret Taylor}

@@

\affil[1]{Department of Wildland Resources and the Ecology Center, Utah State
University, Logan, Utah 84322}
\affil[2]{Department of Ecology and Evolutionary Biology, University of
California, Los Angeles, Los Angeles, California 90095}
\affil[3]{United States Department of Agriculture, Agricultural Research Service,
U.S. Sheep Experiment Station, Dubois, Idaho 83423}

@@

% 
\begin{abstract}
\textbf{Background.} Precipitation is predicted to become more variable
in the western United States, meaning years of above and below average
precipitation will become more common. Periods of extreme precipitation
are major drivers of interannual variability in ecosystem functioning in
water limited communities, but how ecosystems respond to these extremes
over the long-term may shift with precipitation means and variances.
Long-term changes in ecosystem functional response could reflect
compensatory changes in species composition or species reaching
physiological thresholds at extreme precipitation levels.

\textbf{Methods.} We conducted a five year precipitation manipulation
experiment in a sagebrush steppe ecosystem in Idaho, United States. We
used drought and irrigation treatments (approximately 50\%
decrease/increase) to investigate whether ecosystem functional response
remains consistent under sustained high or low precipitation. We
recorded data on aboveground net primary productivity (ANPP), species
abundance, and soil moisture. We fit a generalized linear mixed effects
model to determine if the relationship between ANPP and soil moisture
differed among treatments. We used nonmetric multidimensional scaling to
quantify community composition over the five years.

\textbf{Results.} Ecosystem functional response, defined as the
relationship between soil moisture and ANPP was similar among drought
and control treatments, but the irrigation treatment had a lower slope
than the control treatment. However, ANPP response to availabile soil
moisture was weak and uncertain regardless of treatment, with all slopes
overlapping zero. There was also large spatial variation in ANPP
within-years. Plant community composition was remarkably stable over the
course of the experiment and did not differ among treatments.

\textbf{Discussion.} Despite some evidence that ecosystem functional
response became less sensitive under sustained wet conditions, the
response of ANPP to soil moisture was consistently weak and community
composition was stable. Differences in ecosystem functional responses
across treatments were not related to compensatory shifts at the plant
community level, but instead may reflect the insensitivity of the
dominant species to soil moisture. These species may be successful
precisely because they have evolved life history strategies which buffer
them against precipitation variability.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

@@

At any given site, the relationship between aboveground net primary
productivity (ANPP) and water availability (e.g., soil moisture) can be
characterized by regressing historical observations of ANPP on
observations of soil moisture. The fitted functional response can then
be used to project ANPP under future precipitation regimes (REFERENCE).
A problem with this approach is that it requires extrapolation if future
precipitation falls outside outside the historical range of variability
\citep{Smith2011, Peters2012}. For example, the soil moisture-ANPP
relationship may be linear within the historical range of interannual
variation, but could saturate at higher levels of soil moisture. In
fact, saturating relationships are actually common
\citep{Hsu2012, Gherardi2015a}, perhaps because other resources, like
nitrogen, become more limiting in wet years than dry years. Failure to
accurately estimate the curvature of the soil moisture-ANPP relationship
will lead to over- or underprediction of ANNP under extreme
precipitation \citep{Peters2012}.

Another problem with relying on historical ecosystem functional
responses to predict impacts of altered precipitation regimes is that
these relationships themselves might shift over the long-term. Shifts in
species identities and/or abundances can alter an ecosystem's functional
response to water availability because different species have different
physiological thresholds. \citet{Smith2009} introduced the `Hierarchical
Response Framework' for understanding the interplay of community
composition and ecosystem functioning in response to long-term shifts in
resources. In the near term, ecosystem functioning such as ANPP will
reflect the physiological responses of individual species to the
manipulated resource level. For example, ANPP may decline under
simulated drought because the initial community consisted of
drought-intolerant species \citep{Hoover2014}. Ecosystem functioning may
recover over longer time spans as new species colonize or initial
species reorder in relative abundance. It is also possible that
ecosystem functioning shifts to a new mean state, reflecting the suite
of species in the new community \citep{Knapp2012}.

Experimental manipulations of limiting resources, like precipitation,
offer a route to understanding how ecosystems will respond to resource
levels that fall outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. We propose four alternative scenarios for the effect of
precipitation manipulation on the ecosystem functional response to soil
moisture based on the Hierarchical Response Framework (Fig. 1). We
define `ecosystem functional response' as the relationship between
available soil moisture and ANPP. The four scenarios are based on
possible outcomes at the community (e.g., community composition) and
ecosystem (e.g., soil moisture-ANPP regression) levels.

First, altered precipitation might have no effect on either ecosystem
functional response or community composition (Fig. 1, top left). In this
case, changes in ANPP would be well predicted by the current, observed
soil moisture-ANPP relationship. This corresponds to the early phases of
the Hierarchical Response Framework, where ecosystem response follows
the physiological responses of individual species. Second, the ecosystem
functional response might change while community composition remains the
same (Fig. 1, top right). A saturating soil moisture-ANPP response fits
this scenario, where individual species hit physiological thresholds or
are limited by some other resource. Third, the ecosystem functional
response might be constant but community composition changes (Fig. 1,
bottom left). In this case, changes in species' identities and/or
abundances occur in response to altered precipitation levels and species
more suited to the new conditions compensate for reduced function of
initial species. Fourth, and last, both ecosystem functional response
and community composition could change (Fig. 1, bottom right). New
species, or newly abundant species, with different physiological
responses completely reshape the ecosystem functional response.

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ
\citep{Smith2009, Wilcox2016, Knapp2017}. To determine these time
scales, we need to amass information on how quickly ecosystem functional
responses change in different ecosystems. We also need to understand
whether changes at the ecosystem level are driven by community level
changes or individual level responses.

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (approximately \(\pm50\%\)) and
measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the
relationship between interannual variation in available soil moisture
and interannual variation in ANPP, and if community dynamics underlie
the ecosystem responses. In particular, we are interested in the
consistency of the soil moisture-ANPP relationship among treatments. Is
the relationship steeper under the drought treatment at low soil
moisture? Does the relationship saturate under the irrigation treatment
at high soil moisture? To answer these questions we fit a generalized
linear mixed effects model to test whether the regressions differed
among treatments. We also analyzed community composition over time,
allowing us to place our experimental results within the framework our
scenarios (Fig. 1).

@@

We conducted our precipitation manipulation experiment in a sagebrush
steppe community at the USDA, U.S. Sheep Experiment Station (USSES) near
Dubois, Idaho (44.2\(^{\circ}\) N, 112.1\(^{\circ}\) W), 1500 m above
sea level. The plant community is dominated by the shrub \emph{Artemesia
tripartita} and three perennial bunchgrasses, \emph{Pseudoroegneria
spicata}, \emph{Poa secunda}, and \emph{Hesperostipa comata}. During the
period of our experiment (2011 -- 2015), average mean annual
precipitation was 265 mm year\(\phantom{}^{-1}\) and mean monthly
temperature ranged from -5.2\(^{\circ}\)C in January to
21.8\(^{\circ}\)C in July. Between 1926 and 1932, range scientists at
the USSES established 26 permanent 1 m\(^2\) quadrats to track
vegetation change over time. In 2007, we relocated 14 of the original
quadrats, six of which were inside a large, permanent livestock
exclosure. We use these six plots as control plots (i.e.~ambient
precipitation) in the experiment described below.

In spring 2011, we established 16 new 1 m\(^2\) plots located in the
same exclosure as the six control plots. We avoided areas on steep hill
slopes, areas with greater than 20\% cover of bare rock, and areas with
greater than 10\% cover of the shrubs \emph{Purshia tridentata} and/or
\emph{Amelanchier utahensis}. We established the new plots in pairs and
randomly assigned each plot in a pair to receive a ``drought'' or
``irrigation'' treatment.

@@

Drought and irrigation treatments were designed to decrease and increase
the amount of ambient precipitation by 50\%. To achieve this, we used a
system of rain-out shelters and automatic irrigation
\citep{Gherardi2013}. The rain-out shelters consisted of transparent
acrylic shingles 1-1.5 m above the ground that covered an area of
\(2.5\times2\) m. The shingles intercepted approximately 50\% of
incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on
to the adjacent irrigation plot via two suspended sprinklers. Pumping
was triggered by float switches once water levels reached about 20
liters. We disconnected the irrigation pumps each October and
reconnected them each April. The rain-out shelters remained in place
throughout the year.

We monitored soil moisture in four of the drought-irrigation pairs using
Decagon Devices (Pullman, Washington) 5TM and EC-5 soil moisture
sensors. We installed four sensors around the edges of each 1x1 m census
plot, two at 5 cm soil depth and two at 25 cm soil depth. We also
installed four sensors in areas nearby the four selected plot pairs to
measure ambient soil moisture at the same depths. Soil moisture
measurements were automatically logged every four hours. We coupled this
temporally intensive soil moisture sampling with spatially extensive
readings taken at six points within all 16 plots and associated ambient
measurement areas. These snapshot data were collected on 06-06-2012,
04-29-2015, 05-07-2015, 06-09-2015, and
05-10-2016\footnote{Dates formatted as: mm-dd-yyyy.} using a handheld
EC-5 sensor.

Analyzing the response to experimental treatments was complicated by the
fact that we did not directly monitor soil moisture in each plot on each
day of the experiment. Only a subset of plots were equipped with soil
moisture sensors, and within those plots, one or more of the sensors
frequently failed to collect data. To remedy these problems, and to
produce average daily soil moisture values for the ambient, drought, and
irrigation treatments, we used a statistical model to predict the
average treatment effects on soil moisture during the course of the
experiment.

@@

We then modeled the daily deviation from ambient conditions of the
drought and irrigation treatments using a linear mixed effects model
with independent variables for treatment, season (winter, spring,
summer, fall), rainfall, and all two-way interactions. Rainy days were
defined as any day in which precipitation was recorded and average
temperature was above 3\(^{\circ}\)C. The day immediately following
rainfall was also classified as rainy. We fit the model using the
\texttt{lme4::lmer()} function \citep{Bates2015} in R \citep{R2016},
with random effects for plot group and date. We weighted observations by
the number of unique sensors or spot measurements that were taken in
each plot on that day. We then used the model to predict the average
daily soil moisture in the treated plots based on the average daily
ambient soil moisture. We could only predict soil moisture in the
treated plots on days for which we took at least one ambient soil
moisture measurement. See \citet{Kleinhesselink2017b} for complete
details on our approach to estimating soil moisture.

@@

We estimated aboveground net primary productivity (ANPP) using a
radiometer to relate ground reflectance to plant biomass \citep[see][
for a review]{Byrne2011}. We recorded ground reflectance at four
wavelengths, two associated with red reflectance (626 nm and 652 nm) and
two associated with near-infrared reflectance (875 nm and 859 nm). At
each plot in each year, we took four readings of ground reflectances at
the above wavelengths. We also took readings in 12 (2015), 15 (2012,
2013, 2014), or 16 (2016) calibration plots adjacent to the experimental
site, in which we harvested all aboveground biomass produced in the
current year (we excluded litter and standing dead material), dried it
to a constant weight at 60\(^{\circ}\)C, and weighed it to estimate
ANPP. We harvested at peak biomass each year, typically in late June.

For each plot and year, we averaged the four readings for each
wavelength and then calculated a greenness index based on the same bands
used to calculate NDVI using the MODIS and AVHRR {[}YOU HAVEN'T DEFINED
THESE{]} bands for NDVI. We regressed the greenness index against the
dry biomass weight from the ten calibration plots to convert the
greenness index to ANPP. We fit regressions to a MODIS-based index and
an AVHRR-based index for each year and retained the regression with the
better fit based on \(R^2\) values. We then predicted ANPP using the
best regression equation for each year (Appendix 1).

Species composition data came from two sources: yearly census maps for
each plot made using a pantograph \citep{Hill1920} and yearly counts of
annual species in each plot. From these sources, we determined the
density of all annuals and perennials forbs, the basal cover of
perennial grasses, and the canopy cover of shrubs. We made a large
plot-treatment-year by species matrix, where columns were filled with
either basal cover or density, depending on the measurement made for the
particular species. We standardized the values in each column so we
could directly compare species quantified with different metrics
(density, basal cover, and canopy cover). This puts all abundance values
on the same scale, meaning that common and rare species are weighted
equally. Assuming that rare species will respond to treatments more than
common ones, our approach is biased towards detecting compositional
changes.

@@

\noindent{}where \(\mu_{i}\) is the deterministic prediction from the
regression model for observation \emph{i}, which is associated with plot
\emph{j} and treatment year \emph{t}. \(\boldsymbol{\beta}\) is the
vector of coefficients for the fixed effects in the design matrix
\(\textbf{X}\). Each row of the design matrix represents a single
observation (\(\textbf{x}_i\)) and is a vector with the following
elements: 1 for the intercept, a binary 0 or 1 if the treatment is
``drought'', a binary 0 or 1 if the treatment is ``irrigation'', the
scaled value of VWC, binary ``drought'' value times VWC, and binary
``irrigation'' value times VWC. Thus, our model treats ``control''
observations as the main treatment and then estimates intercept and
slope offsets for the ``drought'' and ``irrigation'' treatments. We use
our model to test the following hypotheses:

@@

These hypotheses are based on evidence that precipitation-ANPP
relationships often saturate with increasing precipitation (REFERENCES).

@@

We used nonmetric multidimensional scaling (NMDS) based on Bray-Curtis
distances to identify temporal changes in community composition among
treatments. We first calculated Bray-Curtis distances among all plots
for each year of the experiment and then extracted those distances for
use in the NMDS. Some values of standardized species' abundances were
negative, which is not allowed for calculating Bray-Curtis distances. We
simply added `2' to each abundance value to ensure all values were
greater than zero. We plotted the first two axes of NMDS scores to see
if community composition overlapped, or not, among treatments in each
year. We used the \texttt{vegan::metaMDS()} function \citep{Oksanen2016}
to calculate Bray-Curtis distances and then to run the NMDS analysis. We
used the \texttt{vegan::adonis()} function \citep{Oksanen2016} to
perform permutational multivariate analysis of variance to test whether
treatment plots formed distinct groupings. To test whether treatment
plots were equally dispersed, or not, we used the
\texttt{vegan::betadisper()} function \citep{Oksanen2016}.

@@

Three of our five study years received below average precipitation (Fig.
2A). Because our treatments are proportional changes in ambient
precipitation, those three years represent a lower magnitude of absolute
difference in precipitation among the treatments. ANPP varied from a
minimum of 74.5 g m\(^{-2}\) in 2014 to a maximum of 237.1 g m\(^{-2}\)
in 2016 when averaged across treatments (Fig. 2C). ANPP was slightly
higher in irrigation plots and slightly lower in drought plots (Fig.
2C), corresponding to estimated soil volumetric water content (VWC)
differences among treatments (Fig. 2B). Such differences in soil VWC
indicate our treatment infrastructure was successful. ANPP was highly
variable across plots within years (Fig. 2C).

@@

Ecosystem response to precipitation extremes depends on the
physiological responses of constituent species and the rate at which
community composition shifts to favor species better able to take
advantage of, or cope with, new resource levels \citep{Smith2009}.
Previous work has shown that community compositional shifts can be both
rapid, on the order of years \citep{Hoover2014}, and slow, on order of
decades \citep{Knapp2012, Wilcox2016}. A lingering question is how the
time scales of ecosystem response and community change vary among
ecosystems. Precipitation manipulation experiments can help answer this
question, especially if they push water availability outside the
historical range of variability for long periods.

The results of our five year experiment in a sagebrush steppe conform to
two of our four predictions, depending on treatment. Neither ecosystem
functional response nor community composition changed under chronic
drought (Fig. 3A, Fig. 4), representin the top left SCENARIO in Fig. 1.
Ecosystem functional response under chronic irrigation was different
from the control treatment, but community composition remained unchanged
(Fig. 3A, Fig. 4), representing the top right SCENARIO in Fig. 1. The
altered ecosystem functional response under irrigation matched our
expectations because we found evidence for a saturating response for the
irrigation treatment (Fig. 3). However, the response of ANPP to soil
moisture in all treatments was consistently weak (Table 1, Fig. A2-2)
and within-year variation of ANPP was large (Fig. 3B).

@@

First, it could be that our drought manipulation was not large enough to
induce a response. That is, a 50\% decrease in any given year may not be
abnormal given our site's historical range of variability
\citep{Knapp2017}. We cannot definitively rule out this possibility, but
we have reason to believe our drought treatment \emph{should} have been
large enough. Using the methods described by \citet{Lemoine2016}, we
calculated the percent reduction and increase of mean growing season
precipitation necessary to reach the 1\% and 99\% extremes of the
historical precipitation regime at our site (Fig. A3-3). The 1\%
quantile of precipitation at our site is 110 mm, a 47\% reduction from
the mean, and the 99\% quantile is 414 mm, a 77\% increase from mean
growing season precipitation (Appendix 3). Thus, our drought treatment
represented extreme precipitation amounts, especially in years where
ambient precipitation was below average (Fig. 2A). The irrigation
treatment may not qualify as extreme, yet that is the treatment where we
did observe an effect (Fig. 3).
